2-the-cubble-pkg

### 2023-05-29 3-the-cubble-pkg.Rmd
update code and rearrange subsection

The \code{cubble} class is an S3 class [@advr] built on the \code{tibble} class, specifically to organize spatio-temporal data. The \code{cubble} class uses an attribute "form",  to arrange the spatial or temporal data components tidily. The form attribute can take a value of either "nested" or "long". The nested \code{cubble} is a subclass of rowwise \code{tibble} (\code{rowwise_df}). It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long \code{cubble} is a subclass of grouped \code{tibble} (\code{grouped_df}), which expands the temporal information into the long form and stores the spatial information in a "spatial" attribute. The main functions in the package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}.

The data \code{climate_flat} is used to illustrate functionality in the \pkg{cubble} package. This is a subset from National Oceanic and Atmospheric Administration (NOAA) [@noaa] Global Historical Climatology Network (GHCN) Daily data. It contains spatial variables, station id, latitude, longitude, elevation, station name, world meteorology organisation id, in additional to daily temporal information,  maximum and minimum temperature values and precipitation records for year 2020. The first five rows of the data are shown below:

```{r}
cubble::climate_flat
```

## Casting into the cubble class {#create}

  - key, index, coords
  - import different data: an expand data, spatial + temporal, netcdf, stars, sftime

@@

```{r echo = TRUE}
cubble_nested <- climate_flat |>
  as_cubble(key = id, index = date, coords = c(long, lat))
cubble_nested
```

```{r echo = TRUE}
cubble_nested |>
  mutate(rain_day = sum(ts$prcp != 0))
```

@@

@@

@@

```{r echo = TRUE}
cubble_long <- cubble_nested |> face_temporal()
cubble_long
```

@@

```{r echo = TRUE}
cubble_back <- cubble_long |> face_spatial()
cubble_back
identical(cubble_nested, cubble_back)
```

@@

```{r echo = TRUE}
cubble_unfold <- cubble_long |> unfold(long, lat)
cubble_unfold
```

@@

```{r echo = TRUE}
climate_flat_ts <- climate_flat |> 
  tsibble::as_tsibble(key = id, index = date)
climate_flat_cb <-  climate_flat_ts |> 
  cubble::as_cubble(coords = c(long, lat))
climate_flat_cb
```

@@

```{r echo = TRUE}
climate_flat_cb |> 
  mutate(fabletools::features(
    ts, tmax, list(tmax_mean = mean, tmax_var = var)
    ))
```

@@

<!-- https://www.jstatsoft.org/style#my-latex-paper-does-not-compile-when-there-is-jss-markup-in-section-titles-what-should-i-do -->

@@

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. The converted sftime object is `r library(cubble); object.size(sftime::st_as_sftime(climate_aus))/object.size(climate_aus)` times the size of the cubble object:

```{r echo = TRUE}
climate_aus %>% sftime::st_as_sftime() %>% object.size()
climate_aus %>% object.size()
```

>>>>

The \code{cubble} class is an S3 class [@advr] built on the \code{tibble} class, specifically to organize spatio-temporal data. The \code{cubble} class uses an attribute "form",  to arrange the spatial or temporal data components tidily. The form attribute can take a value of either "nested" or "long". The nested \code{cubble} is a subclass of rowwise \code{tibble} (\code{rowwise_df}). It arranges each spatial site in a row, and uses list columns to store the feature geometry and the temporal information. The long \code{cubble} is a subclass of grouped \code{tibble} (\code{grouped_df}), which expands the temporal information into the long form and stores the spatial information in a "spatial" attribute.

```{r}
cb_nested <- climate_mel
```

```{r echo = TRUE}
cb_nested
class(cb_nested)
```

```{r}
cb_long <- face_temporal(climate_mel)
```

```{r echo = TRUE}
cb_long
class(cb_long)
```

<!-- The data \code{climate_flat} is used to illustrate functionality in the \pkg{cubble} package. This is a subset from National Oceanic and Atmospheric Administration (NOAA) [@noaa] Global Historical Climatology Network (GHCN) Daily data. It contains spatial variables, station id, latitude, longitude, elevation, station name, world meteorology organisation id, in additional to daily temporal information,  maximum and minimum temperature values and precipitation records for year 2020. The first five rows of the data are shown below: -->

## Creation and coercion {#create}

### Create from separate spatial and temporal table

```{r echo = TRUE}
make_cubble(
  spatial = stations, temporal = meteo,
  key = id, index = date, coords = c(long, lat)
  )
```

```{r echo = TRUE, eval = FALSE}
make_cubble(spatial = stations_sf, temporal = meteo, 
            key = id, index = date)

make_cubble(spatial = stations, temporal = meteo_ts, 
            coords = c(long, lat))

make_cubble(spatial = stations_sf, temporal = meteo_ts)
```

### Coerce from other objects

  - import different data: an expand data, spatial + temporal, netcdf, stars, sftime

@@

```{r echo = TRUE}
climate_flat |> as_cubble(key = id, index = date, coords = c(long, lat))
```

@@

*stars*

*sftimes*

@@

```{r echo = TRUE}
cb_nested |> mutate(rain_day = sum(ts$prcp != 0))
```

@@

```{r echo = TRUE}
(cb_long <- cb_nested |> face_temporal())
```

@@

```{r echo = TRUE}
(cb_back <- cb_long |> face_spatial())
#identical(cb_nested, cb_back)
```

@@

```{r echo = TRUE}
cb_long |> unfold(long, lat)
```

@@

```{r echo = TRUE}
cb <- make_cubble(
  spatial = stations, temporal = meteo_ts, 
  coords = c(long, lat))
```

@@

```{r echo = TRUE}
cb %>% face_temporal() %>% tsibble::has_gaps()
```

@@

```{r echo = TRUE}
cb <- make_cubble(
  spatial = stations_sf, temporal = meteo, 
  key = id, index = date)
class(cb)
#cb %>% sf::st_transform(crs = "EPSG:3857")
```

@@

Consider the climate data used in the \pkg{cubble} package, \code{climate_aus}. It contains daily precipitation, minimum and maximum temperature for 639 weather stations across Australia in 2020. The converted sftime object is `r library(cubble); #object.size(sftime::st_as_sftime(climate_aus))/object.size(climate_aus)` times the size of the cubble object:

```{r echo = TRUE}
#climate_aus %>% sftime::st_as_sftime() %>% object.size()
climate_aus %>% object.size()
```
